A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-08-20 main.tex
2020-08-20 meeting notes/edits

@@

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. The proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ is equal to $||\bX \boldsymbol{\tau}||_F / ||\bZ \bgamma||_F \in [0, 1]$. Since $\bX$ is fixed, this quantity is reflected in $||\boldsymbol{\tau}||_F$, which also quantifies the estimation bias due environmental heterogeneity. In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates.

In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure that can be inferred from $\bX$ will impact the extent to which environmental heterogeneity confounds SNP-phenotype associations.

@@

We begin by comparing the ordinary lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) in various settings and broken down into bias and variance components.  Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

@@

\anna{Obviously these need to be formatted...}\\

@@

\anna{False positive rates}
\begin{verbatim}
   structureX    xi    method      `round(mean(value), 3)` `sd(value)`
 1 1d_linear     0     lasso                         0.038     0.00309
 2 1d_linear     0     lasso_pca10                   0.038     0.00317
 3 1d_linear     0     lmm_ggmix                     0.037     0.00343
 4 1d_linear     0     lmm_lasso                     0.037     0.00320
 5 1d_linear     0     plmm_lasso                    0.038     0.00311
 6 1d_linear     0.8   lasso                         0.039     0.00295
 7 1d_linear     0.8   lasso_pca10                   0.036     0.00307
 8 1d_linear     0.8   lmm_ggmix                     0.035     0.00328
 9 1d_linear     0.8   lmm_lasso                     0.035     0.00327
10 1d_linear     0.8   plmm_lasso                    0.036     0.00298
11 indep_subpops 0     lasso                         0.039     0.00301
12 indep_subpops 0     lasso_pca10                   0.039     0.00310
13 indep_subpops 0     lmm_ggmix                     0.038     0.00367
14 indep_subpops 0     lmm_lasso                     0.038     0.00315
15 indep_subpops 0     plmm_lasso                    0.039     0.00301
16 indep_subpops 0.8   lasso                         0.038     0.00307
17 indep_subpops 0.8   lasso_pca10                   0.032     0.00321
18 indep_subpops 0.8   lmm_ggmix                     0.03      0.00338
19 indep_subpops 0.8   lmm_lasso                     0.03      0.00338
20 indep_subpops 0.8   plmm_lasso                    0.031     0.00318
21 other         0     lasso                         0.031     0.00326
22 other         0     lasso_pca10                   0.031     0.00335
23 other         0     lmm_ggmix                     0.025     0.00451
24 other         0     lmm_lasso                     0.028     0.00358
25 other         0     plmm_lasso                    0.029     0.00335
26 other         0.8   lasso                         0.036     0.00300
27 other         0.8   lasso_pca10                   0.026     0.00338
28 other         0.8   lmm_ggmix                     0.016     0.00348
29 other         0.8   lmm_lasso                     0.02      0.00298
30 other         0.8   plmm_lasso                    0.021     0.00291
\end{verbatim}

@@

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-Lasso-Rakitsch and LMM-Lasso-ggmix for varying $\eta$ levels and sample sizes with independent subpopulation data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}
\anna{add time to this figure? third row...}

>>>>

\anna{REPLACE FIGURES WITH RESULTS FROM PLMM-LASSO ONCE FINISHED}

@@

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ using the ratio $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2 \in [0, 1]$. In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates. In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure that can be inferred from $\bX$, as quantified by $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$, will impact the extent to which environmental heterogeneity confounds SNP-phenotype associations.

@@

We begin by comparing the ordinary lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) in various settings and broken down into bias and variance components.  Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

@@

\anna{Obviously these need to be formatted... just use TP rates - FP not adding much}\\

@@

@@

\anna{add some info/figure on convergence and number of iterations for lmm-lasso vs ggmix, delete the following time figures and references to them}

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-Lasso-Rakitsch and LMM-Lasso-ggmix for varying $\eta$ levels and sample sizes with independent subpopulation data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}
\anna{replace this figure with one that includes time in the third row}
